Stagnating State Funding for Higher Education and its Effect on Faculty at US Universities

### 2023-08-18 text/sections/00-abstract.tex
adjusted writing.

US public universities universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while their states have systematically reduced the funding to these universities relative to enrolment.
This paper connects the changes the faculty composition at US public universities to the fall in the states' funding for their public universities, using a shift-share approach to instrument for state funding and identify the causal effects of stagnating funding for higher education.
A fall in funding of \$1,000 per student leads to univeristies employing 6 more lecturers on average; relative to enrolment, a fall of $10$\% in funding per student decreases the number of assistant professors per student by $1.4$\% and full professors by $1.2$\%, yet increases the number of lecturers per student by $4.4$\%.
Dynamic estimates show that these effects linger multiple years after the initial funding shock.
Analysis of all faculty at Illinois public universities shows that incumbent professors' salaries, promotion rate, and quit rate are all are not affected by funding cuts over 2011-2021, implying that the composition change arose by limiting the hiring of new tenure-track/tenured professors to replace leaving tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and exhibit one possible mechanism for deteriorating student outcomes at US public universities over this same time period.

@@

\vspace{0.1cm}
\noindent
\textbf{JEL Codes:} H72, I23, J45

>>>>

US public universities universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while their states have systematically reduced the funding to these universities relative to enrolment.
This paper connects the changes the faculty composition at US public universities to the fall in the states' funding for their public universities, using a shift-share approach to instrument for state funding and identify the causal effects of stagnating funding for higher education.
A fall in funding of \$1,000 per student leads to univeristies employing 6 more lecturers on average; relative to enrolment, a fall of $10$\% in funding per student decreases the number of assistant professors per student by $1.4$\% and full professors by $1.2$\%, yet increases the number of lecturers per student by $4.4$\%.
Dynamic estimates show that these effects linger multiple years after the initial funding shock.
Analysis of all faculty at Illinois public universities shows that incumbent professors' salaries, promotion rate, and quit rate are all are not affected by funding cuts over 2011-2021, implying that the composition change arose by limiting the hiring of new tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and exhibit one possible mechanism for deteriorating student outcomes at US public universities over this same time period.

@@

\vspace{0.05cm}
\noindent
\textbf{JEL Codes:} H72, I23, J45

### 2023-08-18 text/sections/10-appendix.tex
adjusted writing.

Yearly variation is not observed here, so that only the aggregate level, for 180 universities, can be considered.
\begin{table}[H]
    \singlespacing
    \centering
    \caption{OLS and 2SLS Estimates for University Faculty Hiring, Total for 2011--2020.}
    \makebox[\textwidth][c]{\input{tables/hiring-shock-reg-fte.tex}}
    \label{tab:hiring-shock-reg}
    \justify
    \footnotesize
    \textbf{Note}: 
    This table show the second stage 2SLS estimates of regression specification \eqref{eqn:secondstage}, showing the effect of state funding changes on the number of faculty hires (per student) total for 2011--2021 at US public universities, using the funding shock to instrument for state funding.
    Each observation is a university--year at a public university, where funding data come from IPEDS and faculty count total for 2011--2021 from \citep{wapman2022quantifying,wapman2022zenodo}.
    The panels show the effect of a $1$\% change in state funding per student at the university (total for 2011--2021) on the number of new faculty hires by gender (and all).
    Standard errors are clustered at the university-year level, and university $+$ year fixed effects are included through--out.
\end{table}

>>>>

\newpage
\begin{table}[h!]
    \singlespacing
    \centering
    \caption{OLS and 2SLS Estimates for University Faculty Hiring, Total for 2011--2020.}
    \makebox[\textwidth][c]{\input{tables/hiring-shock-reg-fte.tex}}
    \label{tab:hiring-shock-reg}
    \justify
    \footnotesize
    \textbf{Note}: 
    This table show the second stage 2SLS estimates of regression specification \eqref{eqn:secondstage}, showing the effect of state funding changes on the number of faculty hires (per student) total for 2011--2021 at US public universities, using the funding shock to instrument for state funding.
    Yearly variation in total hires is not observed here, so only the total hires across 2011--2021 for 157 universities, can be considered.
    Each observation is a public university across the years 2011--2021, where data on total funding across 2011--2021 come from IPEDS and faculty count total from \citep{wapman2022quantifying,wapman2022zenodo}.
    The panels show the effect of a $1$\% change in state funding per student at the university (total for 2011--2021) on the number of new faculty hires by gender (and all).
    Standard errors are clustered at the state level, and state fixed effects are included through--out.
\end{table}
